Free Reproduction of Cultural Heritage for Research Purposes: Regulatory Aspects and New Prospects for Project EAGLE

### 2016-01-19 source/29/EAGLE2016_camera_ready.tex
papers revised
minor fixes

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{AUGMENTING THE WORKSPACE OF EPIGRAPHISTS
An interaction design study}

@@

In the fields of epigraphy and archaeology, the areas of digital epigraphy and computational archaeology have benefit
from the use of several of the sensors available in hand-held devices. Crowd-sourcing of photographic data and
geo-spatial information, augmented-reality navigation in archaeological spaces and museums, and 3D scanning of
historical artifacts, using smart phones and tablet computers,, are few of the exemplar applications of hand-held
sensors in epigraphy and archaeology. One common component in all the aforementioned applications is the ability to
record tridimensional data either in the form of geo-spatial coordinates, or in the form of local 3D point coordinates
needed for augmented-reality interaction, or for the construction of triangular meshes of 3D models.

There are several examples in literature that present 3D digitization projects that have been undertaken by museums
including the Epigraphical Museum of Athens (\citet{sullivan_analytical_2011}, \citet{papadaki_accurate_2015}), Museo Arqueológico Nacional de
Madrid \citep{ramirez-sanchez_epigrafidigital:_2014}, Museo Nazionale Romano di Palazzo Altemps \citep{barmpoutis_interactive_2015}, Museo
Geologico Giovanni Capellini di Bologna \citep{abate_valorizzazione_2014}, National Museums Liverpool \citep{cooper_chiswick_2007},
Smithsonian Institution \citep{wachowiak_3d_2009}, and several other museums and institutes (\citet{barsanti_3d_2013}, \citet{landon_petroglyph_2006}, \citet{levoy_digital_2000}).

Several novel methods for scanning, processing, and analyzing 3D models of inscriptions have been developed, including
methods for text extraction from inscriptions (\citet{aswatha_method_2014}, \citet{sullivan_analytical_2011}), accurate 3D scanning of
inscriptions \citep{papadaki_accurate_2015}, visualization of inscriptions \citep{bozia_open-access_2014}, as well as 3D applications for
other archaeological artifacts (\citet{babeu__2011}, \citet{pollefeys_image-based_2001}, \citet{malzbender_polynomial_2001}, \citet{esteban_silhouette_2004}).
Comparative studies of 3D scanning methods for cultural herirage can be found in \citep{pavlidis_methods_2007} and \citep{bohler_3d_2004}.

@@

The recent advances on Natural User Interfaces (NUI) along with their marketing as low-cost general-purpose devices
(smart phones and tablets) have created a nurturing environment for integrating them in cultural heritage applications.
Popular low-cost NUIs, such as touch screens, marker-less position trackers, motion sensors, and head-mounted displays
have been recently studied and employed by museums as mechanisms for multi-sensory virtual experiences (\citet{ujitoko_application_2015}, \citet{soile_accurate_2013}, \citet{ikei_experience_2015}).

@@

@@

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016cameraready-img001.png}
 \includegraphics[width=\columnwidth]{EAGLE2016cameraready-img002.png}
\caption{ Top row: Illustration of interactive manipulation of the virtual lighting by moving the device. The figures
show the corresponding field of normal vectors in the 3D space as computed using the gyroscope of the device. Bottom
row: Demonstration of interactive relighting of a 3D digitized inscription. Different virtual lighting angles reveal
different inscribed details. }
\label{fig:1}
\end{figure}

@@

@@

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016cameraready-img003.png}
\caption{Visualization of same-size boxes using 15 different perspective projections with the same FOV angle and
different cropping parameters. None of the boxes is rotated in the space. }
\label{fig:2}
\end{figure}

@@

@@

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016cameraready-img004.png}
 \includegraphics[width=\columnwidth]{EAGLE2016cameraready-img005.png}
\caption{Top row: Illustration of interactive manipulation of the perspective projection by moving the device and using
the {\textquotedbl}fixed eye{\textquotedbl} assumption. Bottom row: Demonstration of interactive visual inspection of a
3D digitized inscription along with the inscription bearer. The user can view the object from different perspectives,
using natural motions.}
\label{fig:3}
\end{figure}

@@

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016cameraready-img006.png}
\caption{Screenshot of our interactive environment. The user interacts with the 3D object, using touch gestures and
selects one of the regions of interests. This action initiates other data tools such as the image viewer or the edge
filter as shown in this example}
\label{fig:4}
\end{figure}

@@

\noindent In this pilot study, a complete set of natural user interactions was designed based on the physical interactions of
epigraphists with real inscriptions. The proposed interactions utilize the existing sensors in a typical tablet
computer or smart phone in order to interactively relight a digitized inscription and manipulate the user's
perspective, using a set of intuitive gestures that imitate the natural interaction with a physical object. In the
proposed design, the epigraphist can ``hold'' a digital inscription, relight it by
reorienting it as a tangible object, observe it from different perspectives, and finally interact with other modalities
by following a set of 2D touch gestures. The prototype system was developed as part of the Digital Epigraphy and
Archaeology (DEA) project using the open-source library VisiNeat for 3D visualization and interaction and is compatible
with iOS, Android, and Microsoft RT tablet and smart phone devices. The interface is available through the web-site of
the project: http://www.digitalepigraphy.org

>>>>

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Augmenting the Workspace of Epigraphists.
An interaction design study}

@@

In the fields of epigraphy and archaeology, the areas of digital epigraphy and computational archaeology have benefit
from the use of several of the sensors available in handheld devices. Crowd-sourcing of photographic data and
geo-spatial information, augmented-reality navigation in archaeological spaces and museums, and 3D scanning of
historical artifacts, using smart phones and tablet computers, are few of the exemplar applications of handheld
sensors in epigraphy and archaeology. One common component in all the aforementioned applications is the ability to
record tridimensional data either in the form of geo-spatial coordinates, or in the form of local 3D point coordinates
needed for augmented-reality interaction, or for the construction of triangular meshes of 3D models.

There are several examples in literature that present 3D digitization projects that have been undertaken by museums
including the Epigraphical Museum of Athens \citep{sullivan_analytical_2011, papadaki_accurate_2015}, Museo Arqueológico Nacional de
Madrid \citep{ramirez-sanchez_epigrafidigital:_2014}, Museo Nazionale Romano di Palazzo Altemps \citep{barmpoutis_interactive_2015}, Museo
Geologico Giovanni Capellini di Bologna \citep{abate_valorizzazione_2014}, National Museums Liverpool \citep{cooper_chiswick_2007},
Smithsonian Institution \citep{wachowiak_3d_2009}, and several other museums and institutes \citep{barsanti_3d_2013, landon_petroglyph_2006, levoy_digital_2000}.

Several novel methods for scanning, processing, and analyzing 3D models of inscriptions have been developed, including
methods for text extraction from inscriptions \citep{aswatha_method_2014, sullivan_analytical_2011}, accurate 3D scanning of
inscriptions \citep{papadaki_accurate_2015}, visualization of inscriptions \citep{bozia_open-access_2014}, as well as 3D applications for
other archaeological artifacts \citep{babeu__2011, pollefeys_image-based_2001, malzbender_polynomial_2001, esteban_silhouette_2004}.
Comparative studies of 3D scanning methods for cultural herirage can be found in \citep{pavlidis_methods_2007} and \citep{bohler_3d_2004}.

@@

The recent advances on Natural User Interfaces (NUI) along with their marketing as low-cost general-purpose devices
(smart phones and tablets) have created a nurturing environment for integrating them in cultural heritage applications.
Popular low-cost NUIs, such as touch screens, marker-less position trackers, motion sensors, and head-mounted displays
have been recently studied and employed by museums as mechanisms for multi-sensory virtual experiences \citep{ujitoko_application_2015, soile_accurate_2013, ikei_experience_2015}.

@@

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016cameraready-img001.png}
 \includegraphics[width=\columnwidth]{EAGLE2016cameraready-img002.png}
\caption{ Top row: Illustration of interactive manipulation of the virtual lighting by moving the device. The figures
show the corresponding field of normal vectors in the 3D space as computed using the gyroscope of the device. Bottom
row: Demonstration of interactive relighting of a 3D digitized inscription. Different virtual lighting angles reveal
different inscribed details. }
\label{fig:1}
\end{figure}

@@

@@

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016cameraready-img003.png}
\caption{Visualization of same-size boxes using 15 different perspective projections with the same FOV angle and
different cropping parameters. None of the boxes is rotated in the space. }
\label{fig:2}
\end{figure}

@@

@@

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016cameraready-img004.png}
 \includegraphics[width=\columnwidth]{EAGLE2016cameraready-img005.png}
\caption{Top row: Illustration of interactive manipulation of the perspective projection by moving the device and using
the {\textquotedbl}fixed eye{\textquotedbl} assumption. Bottom row: Demonstration of interactive visual inspection of a
3D digitized inscription along with the inscription bearer. The user can view the object from different perspectives,
using natural motions.}
\label{fig:3}
\end{figure}

@@

@@

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016cameraready-img006.png}
\caption{Screenshot of our interactive environment. The user interacts with the 3D object, using touch gestures and
selects one of the regions of interests. This action initiates other data tools such as the image viewer or the edge
filter as shown in this example}
\label{fig:4}
\end{figure}

@@

\noindent In this pilot study, a complete set of natural user interactions was designed based on the physical interactions of
epigraphists with real inscriptions. The proposed interactions utilize the existing sensors in a typical tablet
computer or smart phone in order to interactively relight a digitized inscription and manipulate the user's
perspective, using a set of intuitive gestures that imitate the natural interaction with a physical object. In the
proposed design, the epigraphist can ``hold'' a digital inscription, relight it by
reorienting it as a tangible object, observe it from different perspectives, and finally interact with other modalities
by following a set of 2D touch gestures. The prototype system was developed as part of the Digital Epigraphy and
Archaeology (DEA) project using the open-source library VisiNeat for 3D visualization and interaction and is compatible
with iOS, Android, and Microsoft RT tablet and smart phone devices. The interface is available through the web-site of
the project: \url{http://www.digitalepigraphy.org}

### 2016-01-19 source/38/EAGLE2016_Final_Paper Modolo.tex
papers revised
minor fixes

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{FREE REPRODUCTION OF CULTURAL HERITAGE FOR RESEARCH PURPOSES: REGULATORY ASPECTS AND NEW PROSPECTS FOR PROJECT `EAGLE'}
\author[free]{Mirco Modolo\corref{first}}
\address[free]{Free lance archaeologist and archivist}
\cortext[first]{Corresponding author. Email: mircomodolo@gmail.com}
\date{2015-11-20}
\begin{document}

\maketitle
\begin{abstract}
This paper traces the genesis and activity of the ``Fotografie libere per i Beni Culturali'' movement promoting free
remote reproduction, and reuse for scientific purposes, of manuscripts and antique volumes conserved in Italian
archives and libraries. Simultaneously, evidence highlights the advantages of deregulating cultural heritage images for
historical research, in particular for the study of antique epigraphy developed by the EAGLE project.
\end{abstract}
\keywords{Archivi, Biblioteche, Manoscritti, Riproduzioni, Creative Commons}

@@

On the first of June 2014, the `Art Bonus' decree came into effect, sanctioning the free reproduction of all cultural
heritage for scientific purposes (even reproduction carried out at a distance). It represented a new first for
historical research\footnote{Decreto-legge 31 maggio 2014 n. 83, art. 12.}: numerous archives and libraries opened
and made free of charge the reproduction of manuscripts and historical volumes with a user’s camera. This provided a
clear advantage for historical research, particularly epigraphy, which considers antique manuscripts from the 15th and
19th centuries an important source for the study of Latin and Greek inscriptions.

As a direct consequence of the Art Bonus decree, some institutions, including the State Archives of Florence (Archivio
di Stato di Firenze), allowed users to make digital copies of records in consultation. Unfortunately, just one month
later, a restrictive amendment modified the original text of the law, explicitly excluding printed books, manuscripts
and archival documents from the liberalisation\footnote{Legge 29 luglio 2014, n. 106}. Many scholars’ surprise was
overtaken by irritation, driven by an apparent paradox: museum photography remained free for the tourist, yet the
scholar engaged in research activity in archives and libraries would not reap the benefits of
liberalisation\footnote{\citet{modolo_il_2014}; \citet{brugnoli_ancora_2014}. }.

\begin{enumerate}
\item Current state of the problem
\end{enumerate}
Excluding bibliographic and archival records from liberalisation has effectively restored the previous regime, still in
force today: taxing images taken with a personal digital device (in the institutions still allowing this freedom), or
an outright ban on use of a personal device – requiring the paid use of a photographic service. Rates for using a
personal device are the most varied (on average 3 Euro per unit), while some archives require a rental fee for making
reproductions on a personal device in a specialised room. Thus an extreme variety of tariffs emerges, commoditising
research to the detriment of those who contribute to enhancing documentary heritage through their own study.
Furthermore this is in direct contrast with the Italian Constitution which, according to articles 33 and 9, requires
the Republic not only to guarantee free research, but also to actively promote it. An effective ``tax'' on reproduction
with a personal device becomes a research obstacle; it is absurd to expect a fee for a service provided autonomously by
the researcher himself.

@@

The protection of archival material is one of the principal arguments against free reproduction. Indeed archivists at
the State Archives of Florence (Archivio di Stato di Firenze) directly noted the advantages of free reproduction in the
month ``Art Bonus'' came into force (prior to the approval of the restrictive amendment): one could observe not only that
the images did not damage the archival material, but rather, being subject to less movement, actually contributed to
its conservation\footnote{\url{https://fotoliberebbcc.wordpress.com/2015/06/06/4724/}}.

If the remainder of the problem is indeed conservation, it is difficult to understand why this might be considered a
problem only in Italy: the policies of the National Archives of the United Kingdom and the French Archives Nationales
already foresee free reproduction with a user’s personal device. In recent months the British Library has spread across
the web a short video demonstration in order to instruct the users in the proper handling of the codex manuscript.
Specifically, the video prompts the user to employ simple lead cordons to hold open the pages of a codex without
compressing the back of the volume and risking damaging the
binding\footnote{\url{http://www.bl.uk/reshelp/inrrooms/stp/copy/selfsrvcopy/book\_photography\_video.mp4}.}.

@@

The application of tariffs for reproduction and reuse of cultural heritage images appears, even more than a means of
generating profits, a pretext designed to discourage reproduction of manuscripts and ancient volumes and to minimise
the risk of abuse associated with using the images for profit-making activity, in a way which might `damage' the public
treasury\footnote{\url{http://www.bianchibandinelli.it/2015/05/25/fotografie-libere-un-comunicato-dellassociazione-bianchi-bandinelli/}}.

@@

To meet scholars’ demands, in September 2014 researchers spontaneously formed a movement of ideas, ``Fotografie libere
per i Beni Culturali''. The group launched a petition asking Minister of Culture on. Dario Franceschini for free
reproduction in archives and libraries, renewing the original spirit of the ``Art Bonus''
decree\footnote{\url{https://fotoliberebbcc.wordpress.com/category/adesioni-e-contatti/}}. It was an unprecedented
initiative given not only the number of signatures (more than 4200), but also the quality of subscribers: a chorus
comprising the highest representatives from each of the historical-humanitarian disciplines from around Europe:
associations, university docents, researchers, students, and simple fans of local history, along with directors of
state archives and functionaries of the ministry
itself\footnote{\url{http://www.ilgiornaledellarte.com/articoli/2015/4/123892.html}.}.

Moreover, the movement produced a proposal to amend article 108 of the Codice dei Beni Culturali regulating the
reproduction of cultural heritage (\url{https://fotoliberebbcc.wordpress.com/category/la-nostra-proposta/}), to eliminate the
exclusion of printed books and documents from the liberalisation archive, hence complying with copyright and personal
privacy for archival documents.

@@

Free photography means the ability to use one’s own smartphone or camera in archives and libraries during consultation,
without the need to request prior written approval (for purposes other than profit). A `cultural’ exception should be
provided for a particular form of profit, scientific publishing, given the belief that research is meaningful only if
it can be disseminated in the broadest way\footnote{Circolare Mibact 21/2005.}. ``Fotografie libere per i Beni
Culturali'' in fact proposes to make free the publication of images of cultural heritage in scientific texts in
circulation and of a limited cover price (below 77 Euro and 2000 copies) with the sole duty to always specify the name
of the library/archive of provenance, and to provide a copy of the publication. In such cases the official request for
authorisation by ordinary post may be substituted by a simple communication online at the institute with the intent to
publish. For the ``greater'' publications, other than the usual indicated above, and in general for moneymaking
activities, the concession regime will remain in force, requiring both a formal authorisation request as well as a
payment for the canon of use (royalties).

@@

Free reuse of the images, at least for cultural purposes, has obvious positive effects on the liberal circulation of
scientific content, and also represents an important way to improve the visibility of institutions and their
collections. Museums, archives and libraries should therefore be seen as active and dynamic centres of cultural
promotion, more so than bureaucratic offices for the static conservation of cultural heritage\footnote{Cfr. on the
Creative Commons licences in archaeology: \citet{brugnoli_fotografia_2013}; \citet{gualandi}.}.

@@

The effort underlying the collection of thousands of signatures has already led to encouraging openings. A recent
parliamentary question to the minister on whether to restore the original sprit of the ``Art Bonus''
decree\footnote{\url{http://www.senato.it/japp/bgt/showdoc/showText?tipodoc=Sindisp\&leg=17\&id=914146}}, and the
comparable availability shown by the Ministry of Culture, are in fact events which inspire cautious optimism –
primarily because they mark a return to this topic in parliament after one year. It begins a course that will hopefully
soon arrive at the modification of article 108 of the ``Codice dei Beni Culturali'' regulating reproduction and the
release of an appropriate circular which should affirm, without ambiguity:

@@

\item the free reuse of images of cultural heritage for purposes outside of financial gain, or free publication online
and in print within the limits mentioned above\footnote{Cfr. n. 10.}.

>>>>

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Free Reproduction of Cultural Heritage for Research Purposes:
Regulatory Aspects and New Prospects for Project EAGLE}
\author[free]{Mirco Modolo\corref{first}}
\address[free]{Free lance archaeologist and archivist}
\cortext[first]{Corresponding author. Email: mircomodolo@gmail.com}
\date{2015-11-20}
\begin{document}

\maketitle
\begin{abstract}
This paper traces the genesis and activity of the ``Fotografie libere per i Beni Culturali'' movement promoting free
remote reproduction, and reuse for scientific purposes, of manuscripts and antique volumes conserved in Italian
archives and libraries. Simultaneously, evidence highlights the advantages of deregulating cultural heritage images for
historical research, in particular for the study of antique epigraphy developed by the EAGLE project.
\end{abstract}
\keywords{Archives, Libraries, Manuscripts, Reproductions, Creative Commons}

@@

On the first of June 2014, the `Art Bonus' decree came into effect, sanctioning the free reproduction of all cultural
heritage for scientific purposes (even reproduction carried out at a distance). It represented a new first for
historical research:\footnote{Decreto-legge 31 maggio 2014 n. 83, art. 12.} numerous archives and libraries opened
and made free of charge the reproduction of manuscripts and historical volumes with a user’s camera. This provided a
clear advantage for historical research, particularly epigraphy, which considers antique manuscripts from the 15th and
19th centuries an important source for the study of Latin and Greek inscriptions.

As a direct consequence of the Art Bonus decree, some institutions, including the State Archives of Florence (Archivio
di Stato di Firenze), allowed users to make digital copies of records in consultation. Unfortunately, just one month
later, a restrictive amendment modified the original text of the law, explicitly excluding printed books, manuscripts
and archival documents from the liberalisation.\footnote{Legge 29 luglio 2014, n. 106} Many scholars’ surprise was
overtaken by irritation, driven by an apparent paradox: museum photography remained free for the tourist, yet the
scholar engaged in research activity in archives and libraries would not reap the benefits of
liberalisation.\footnote{\citet{modolo_il_2014}; \citet{brugnoli_ancora_2014}. }

\section{Current state of the problem}

Excluding bibliographic and archival records from liberalisation has effectively restored the previous regime, still in
force today: taxing images taken with a personal digital device (in the institutions still allowing this freedom), or
an outright ban on use of a personal device – requiring the paid use of a photographic service. Rates for using a
personal device are the most varied (on average 3 Euro per unit), while some archives require a rental fee for making
reproductions on a personal device in a specialised room. Thus an extreme variety of tariffs emerges, commoditising
research to the detriment of those who contribute to enhancing documentary heritage through their own study.
Furthermore this is in direct contrast with the Italian Constitution which, according to articles 33 and 9, requires
the Republic not only to guarantee free research, but also to actively promote it. An effective ``tax'' on reproduction
with a personal device becomes a research obstacle; it is absurd to expect a fee for a service provided autonomously by
the researcher himself.

@@

The protection of archival material is one of the principal arguments against free reproduction. Indeed archivists at
the State Archives of Florence (Archivio di Stato di Firenze) directly noted the advantages of free reproduction in the
month ``Art Bonus'' came into force (prior to the approval of the restrictive amendment): one could observe not only that
the images did not damage the archival material, but rather, being subject to less movement, actually contributed to
its conservation.\footnote{\url{https://fotoliberebbcc.wordpress.com/2015/06/06/4724/}}

If the remainder of the problem is indeed conservation, it is difficult to understand why this might be considered a
problem only in Italy: the policies of the National Archives of the United Kingdom and the French Archives Nationales
already foresee free reproduction with a user’s personal device. In recent months the British Library has spread across
the web a short video demonstration in order to instruct the users in the proper handling of the codex manuscript.
Specifically, the video prompts the user to employ simple lead cordons to hold open the pages of a codex without
compressing the back of the volume and risking damaging the
binding.\footnote{\url{http://www.bl.uk/reshelp/inrrooms/stp/copy/selfsrvcopy/book\_photography\_video.mp4}.}

@@

The application of tariffs for reproduction and reuse of cultural heritage images appears, even more than a means of
generating profits, a pretext designed to discourage reproduction of manuscripts and ancient volumes and to minimise
the risk of abuse associated with using the images for profit-making activity, in a way which might `damage' the public
treasury.\footnote{\url{http://www.bianchibandinelli.it/2015/05/25/fotografie-libere-un-comunicato-dellassociazione-bianchi-bandinelli/}}

@@

To meet scholars’ demands, in September 2014 researchers spontaneously formed a movement of ideas, ``Fotografie libere
per i Beni Culturali''. The group launched a petition asking Minister of Culture on. Dario Franceschini for free
reproduction in archives and libraries, renewing the original spirit of the ``Art Bonus''
decree.\footnote{\url{https://fotoliberebbcc.wordpress.com/category/adesioni-e-contatti/}} It was an unprecedented
initiative given not only the number of signatures (more than 4200), but also the quality of subscribers: a chorus
comprising the highest representatives from each of the historical-humanitarian disciplines from around Europe:
associations, university docents, researchers, students, and simple fans of local history, along with directors of
state archives and functionaries of the ministry
itself.\footnote{\url{http://www.ilgiornaledellarte.com/articoli/2015/4/123892.html}.}

Moreover, the movement produced a proposal to amend article 108 of the Codice dei Beni Culturali regulating the
reproduction of cultural heritage,\footnote{\url{https://fotoliberebbcc.wordpress.com/category/la-nostra-proposta/}} to eliminate the
exclusion of printed books and documents from the liberalisation archive, hence complying with copyright and personal
privacy for archival documents.

@@

Free photography means the ability to use one’s own smartphone or camera in archives and libraries during consultation,
without the need to request prior written approval (for purposes other than profit). A `cultural’ exception should be
provided for a particular form of profit, scientific publishing, given the belief that research is meaningful only if
it can be disseminated in the broadest way.\footnote{Circolare Mibact 21/2005.} ``Fotografie libere per i Beni
Culturali'' in fact proposes to make free the publication of images of cultural heritage in scientific texts in
circulation and of a limited cover price (below 77 Euro and 2000 copies) with the sole duty to always specify the name
of the library/archive of provenance, and to provide a copy of the publication. In such cases the official request for
authorisation by ordinary post may be substituted by a simple communication online at the institute with the intent to
publish. For the ``greater'' publications, other than the usual indicated above, and in general for moneymaking
activities, the concession regime will remain in force, requiring both a formal authorisation request as well as a
payment for the canon of use (royalties).

@@

Free reuse of the images, at least for cultural purposes, has obvious positive effects on the liberal circulation of
scientific content, and also represents an important way to improve the visibility of institutions and their
collections. Museums, archives and libraries should therefore be seen as active and dynamic centres of cultural
promotion, more so than bureaucratic offices for the static conservation of cultural heritage.\footnote{Cfr. on the
Creative Commons licences in archaeology: \citet{brugnoli_fotografia_2013}; \citet{gualandi}.}

@@

The effort underlying the collection of thousands of signatures has already led to encouraging openings. A recent
parliamentary question to the minister on whether to restore the original sprit of the ``Art Bonus''
decree,\footnote{\url{http://www.senato.it/japp/bgt/showdoc/showText?tipodoc=Sindisp\&leg=17\&id=914146}} and the
comparable availability shown by the Ministry of Culture, are in fact events which inspire cautious optimism –
primarily because they mark a return to this topic in parliament after one year. It begins a course that will hopefully
soon arrive at the modification of article 108 of the ``Codice dei Beni Culturali'' regulating reproduction and the
release of an appropriate circular which should affirm, without ambiguity:

@@

\item the free reuse of images of cultural heritage for purposes outside of financial gain, or free publication online
and in print within the limits mentioned above.\footnote{Cfr. n. 10.}
